---
title: "Usage of functions"
author:
  - name: Mariano Ruz Jurado
    affiliation:
    - Goethe University
output: 
  BiocStyle::html_document:
    self_contained: true
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
package: "`r pkg_ver('DOtools')`"
vignette: >
  %\VignetteIndexEntry{Usage of functions}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: inline
---

```{r chunk_setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

```{r vignette_setup, echo=FALSE, message=FALSE, warning = FALSE}
# Track time spent on making the vignette.
start_time <- Sys.time()
pkg <- function(x) {
  paste0("[*", x, "*](https://cran.r-project.org/package=", x, ")")
}
pypkg <- function(x) {
  paste0("[*", x, "*](https://pypi.org/project/", x, "/)")
}
```

In this tutorial we compile examples on how to use the different functions available in the R package. You can refer to previous tutorial to see how to use CellBender and perform quality control and basic downstream analysis of sc/snRNA-seq. We are going to use the object generated in the section: Quality control of sc/snRNA-seq.

## Libraries & Set-Up
```{r load_library, message=FALSE, echo=FALSE}
library(DOtools)

DO.PyEnv()
reticulate::use_python("~/.venv/DOtools/bin/python")
Seu_obj <- readRDS(system.file("extdata", "sc_data.rds", package = "DOtools"))
```

## Dotplot

The `DO.Dotplot` function generates a plot in a similar manner as the Dotplot function implemented in `r pypkg("scanpy")` but with additional functionalities, e.g. the visualisation of 3 variables at the same time. For example, we can visualise the expression of a gene across celltypes and conditions:

```{r dotplot, fig.width=5, fig.height=5}
DO.Dotplot(Seu_object = Seu_obj,
           group.by.x = "condition",
           group.by.y = "annotation",
           Feature = "NKG7")
```

## UMAP

The `DO.UMAP` function facilitatesthe visualisation of UMAP embeddings since it combines the ability to show cell groupings and gene expression.

```{r UMAP, fig.width=12, fig.height=8, warning = FALSE}

DO.UMAP(Seu_object = Seu_obj,
        group.by = "annotation",
        split.by = "condition",
        FeaturePlot = TRUE,
        features = "NKG7",
        legend.position = "right")
```

## Cell composition changes

As shown in the tutorial, we can also test for significant changes in cell population.

```{r CellComposition, fig.width=6, fig.height=5, warning = FALSE}

DO.CellComposition(Seu_object = Seu_obj,
                   assay_normalized = "RNA",
                   cluster_column = "annotation",
                   sample_column = "orig.ident",
                   condition_column = "condition",
                   transform_method = "arcsin",
                   n_reps = 3)

```

## Expression of genes and continuous metadata

We can visualize the average expression of a gene in a cell type or continuous metadata information across condition with violinplots, barplots and boxplots. Additionally, we can test for significance.

```{r Violin, fig.width=8, fig.height=5.5, warning = FALSE}
p <- DO.VlnPlot(Seu_obj,
                Feature = "NKG7",
                group.by = "condition",
                group.by.2 = "annotation",
                ctrl.condition = "healthy")
```

```{r Bar, fig.width=5, fig.height=6, warning = FALSE}
Seu_obj_NK <- DO.Subset(Seu_obj, ident = "annotation", ident_name = "NK")
DO.BarplotWilcox(Seu_obj_NK,
                 group.by = "condition",
                 ctrl.condition = "healthy",
                 Feature = "NKG7",
                 x_label_rotation = 0)
```

```{r Box, fig.width=5, fig.height=6, warning = FALSE}
set.seed(123)
Seu_obj$rdm_sample <- sample(rep(c("A", "B", "C"), length.out = ncol(Seu_obj)))
DO.BoxPlot(Seu_obj,
           group.by = "rdm_sample",
           ctrl.condition = "A",
           Feature = "nCount_RNA",
           step_mod = 100,
           plot_sample = FALSE)
```


```{r Heat map, out.width="100%", fig.align="center", fig.width=10, fig.height=8, warning = FALSE}
path_file <- tempfile("dotools_plots_")
dir.create(path_file, recursive = TRUE, showWarnings = FALSE)
DO.Heatmap(Seu_obj,
           features = rownames(Seu_obj)[1:10],
           xticks_rotation=45,
           path = path_file,
           stats_x_size = 20)
Heatmap_plot <- list.files(path = path_file, pattern = "Heatmap*\\.svg$", full.names = TRUE, recursive = TRUE)
plot(magick::image_read_svg(Heatmap_plot))

```


# Session information

```{r session_info, echo=FALSE}
options(width = 120)
sessioninfo::session_info()
```
